Expert-In-The-Loop Causal Discovery: \\ Iterative Model Refinement Using Expert Knowledge

### 2025-02-11 paper/main.tex
UPdates plot and caption
Updates

\section{Background}
\label{sec:background}
We denote random variables using uppercase letters like $X$, and a set of
random variables by $ \bm{X} = \{X_1, \cdots, X_k\} $, where $ \rvert \bm{X}
\rvert = k $ is the number of variables in the set. We denote the
covariance between variables $ X $ and $ Y $ as $ \mathrm{cov}(X, Y) $, a 
covariance matrix as $ \Sigma $, and the entry corresponding to the covariance
between $ X $ and $ Y $ as $ \Sigma_{XY} $. We consider causal DAGs that contain 
only observed variables. A DAG $ G = (V, E) $ is an acyclic directed graph
whose nodes $ V $ corresponding to random variables and whose edges $ E $ represent
direct causal relationships \cite{Pearl2009}. For an edge $e=X \to Y$, 
$\textrm{reverse}(e)=Y \to X$. The set of parents of $ X $ in $ G $
is denoted as $ \textrm{Pa}_G(X) $, and its ancestors and descendants as $
\textrm{An}_G(X) $ and $ \textrm{De}_G(X) $, respectively; we use the convention
that $X \notin \textrm{An}_G(X)$ and $X \notin \textrm{De}_G(X)$. We denote a path, $
\pi(X, Y) = \{ X, V_0, V_1, \cdots, V_k, Y \} $ between $ X $ and $ Y $ in $ G
$ where consecutive pairs of variables in $ \pi $ are connected by an
edge. We say $ X $ and $ Y $ are d-connected in $ G $ given a conditioning set
$ \bm{Z} \subseteq V - \{X, Y\} $, if for a path $ \pi(X, Y) $ (i) for every
collider structure ($ V_1 \rightarrow V_2 \leftarrow V_3 $) on $ \pi(X, Y) $, $
\bm{Z} \cap \{ V_2, \textrm{De}_G(V_2) \} \ne \emptyset $, and (ii) for every
non-collider structure ($ V_1 \rightarrow V_2 \rightarrow V_3 $, $ V_1
\leftarrow V_2 \rightarrow V_3 $), $ V_2 \not \in \bm{Z} $. $ X $ and $ Y $ are
d-separated given $ \bm{Z} $ if no d-connecting path exist between them.

@@

\paragraph{Both $ X $ and $ Y $ are continuous: }
When both $ X $ and $ Y $ are continuous, we can use a (partial) correlation
test for CI testing and Pearson's correlation coefficient can be used as the
effect size. When $ \bm{Z} = \emptyset $, the correlation coefficient is
defined as: $ r_{X, Y} = \mathrm{cov}(X, Y) / (\sigma_X \sigma_Y) $. When $
\bm{Z} \neq \emptyset $, the partial correlation coefficient can be used instead.
This is estimated by fitting two regression models $ E_X: X \sim \bm{Z} $ and $ E_Y: Y
\sim \bm{Z} $, calculating the residuals $ R_X = X - E_X(\bm{Z}) $ and $ R_Y =
Y - E_Y(\bm{Y}) $, and then computing the correlation coefficient between the
residuals: $r_{X, Y \rvert \bm{Z}} = r_{R_X, R_Y}$

@@

\begin{figure}[t!]
	\centering
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics{../code/fig3/shd_ribbon.pdf}
		\caption{}
	\end{subfigure}
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics{../code/fig3/sid_ribbon.pdf}
		\caption{}
	\end{subfigure}
	\caption{Comparison of PC, Hill-Climb Search, and GES algorithms against
		\textsc{ExpertInLoop} algorithm. As automated algorithms only
		recover the CPDAG, we use the best and worst scoring
		orientation of the CPDAG to get the range. We test
		\textsc{ExpertInLoop} with varying values of expert accuracy, $ \alpha = \{0.1, 0.3, 0.5, 0.7, 0.9\} $. The corresponding
		$\alpha_{\textrm{eff}} $ is shown in the plot.}
	\label{fig:shd_sid}
\end{figure}

>>>>

\section{Background}
\label{sec:background}
We denote random variables using uppercase letters like $X$, and a set of
random variables by $ \bm{X} = \{X_1, \cdots, X_k\} $, where $ \rvert \bm{X}
\rvert = k $ is the number of variables in the set. We denote the standard
deviation of $ X $ as $ \sigma_X $, the covariance between variables $ X $ and
$ Y $ as $ \mathrm{cov}(X, Y) $, a covariance matrix as $ \Sigma $, and the
entry corresponding to the covariance between $ X $ and $ Y $ as $ \Sigma_{XY}
$. We consider causal DAGs that contain only observed variables. A DAG $ G =
(V, E) $ is an acyclic directed graph whose nodes $ V $ corresponding to random
variables and whose edges $ E $ represent direct causal relationships
\cite{Pearl2009}. For an edge $e=X \to Y$, $\textrm{reverse}(e)=Y \to X$. The
set of parents of $ X $ in $ G $ is denoted as $ \textrm{Pa}_G(X) $, and its
ancestors and descendants as $ \textrm{An}_G(X) $ and $ \textrm{De}_G(X) $,
respectively; we use the convention that $X \notin \textrm{An}_G(X)$ and $X
\notin \textrm{De}_G(X)$. We denote a path, $ \pi(X, Y) = \{ X, V_0, V_1,
\cdots, V_k, Y \} $ between $ X $ and $ Y $ in $ G $ where consecutive pairs of
variables in $ \pi $ are connected by an edge. We say $ X $ and $ Y $ are
d-connected in $ G $ given a conditioning set $ \bm{Z} \subseteq V - \{X, Y\}
$, if for a path $ \pi(X, Y) $ (i) for every collider structure ($ V_1
\rightarrow V_2 \leftarrow V_3 $) on $ \pi(X, Y) $, $ \bm{Z} \cap \{ V_2,
\textrm{De}_G(V_2) \} \ne \emptyset $, and (ii) for every non-collider
structure ($ V_1 \rightarrow V_2 \rightarrow V_3 $, $ V_1 \leftarrow V_2
\rightarrow V_3 $), $ V_2 \not \in \bm{Z} $. $ X $ and $ Y $ are d-separated
given $ \bm{Z} $ if no d-connecting path exist between them.

@@

\paragraph{Both $ X $ and $ Y $ are continuous: }
When both $ X $ and $ Y $ are continuous, we can use a (partial) correlation
test for CI testing and Pearson's correlation coefficient can be used as the
effect size. When $ \bm{Z} = \emptyset $, the correlation coefficient is
defined as: $ r_{X, Y} = \mathrm{cov}(X, Y) / (\sigma_X \sigma_Y) $. When $
\bm{Z} \neq \emptyset $, the partial correlation coefficient can be used instead.
This is estimated by fitting two regression models $ E_X: X \sim \bm{Z} $ and $ E_Y: Y
\sim \bm{Z} $, calculating the residuals $ R_X = X - E_X(\bm{Z}) $ and $ R_Y =
Y - E_Y(\bm{Y}) $, and computing the Pearson's correlation coefficient between the
residuals: $r_{X, Y \rvert \bm{Z}} = r_{R_X, R_Y}$

@@

\begin{figure}[t!]
	\centering
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics{../code/fig3/shd_ribbon.pdf}
		\caption{}
	\end{subfigure}
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics{../code/fig3/sid_ribbon.pdf}
		\caption{}
	\end{subfigure}
	\caption{Comparison of PC, Hill-Climb Search, and GES algorithms against
		\textsc{ExpertInLoop} algorithm. As automated algorithms only
		recover the CPDAG, we use the best and worst scoring
		orientation of the CPDAG to get the range. We test
		\textsc{ExpertInLoop} with varying values of expert accuracy, $ \alpha = \{0.1, 0.3, 0.5, 0.7, 0.9\} $. The corresponding
	$\alpha_{\textrm{eff}} $ is shown in the plot. Expert shading: mean $\pm$ standard error. Others shading: [mean min. - standard error min., mean max. + standard error max.]}
	\label{fig:shd_sid}
\end{figure}
